Towards a Verified Tableau Prover for a Quantifier-Free Fragment of Set Theory

### 2023-05-23 paper/CADE/paper.tex
Realise reviewer comments

\usepackage{appendix}

@@

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is a quantifier-free fragment of set theory.
  We formalise its syntax and semantics as well as a sound and complete tableau calculus for it.
  We also provide an executable specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
  Furthermore, we extend the calculus with a lightweight type system that paves the way for an integration of the procedure into Isabelle/HOL.
  \keywords{Decision procedures \and Tableau \and Proof assistants \and Set theory}
\end{abstract}

@@

\subsection{Contributions}
We present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
In order to obtain an executable procedure, we also give a naive refinement to a more concrete specification from which code can be generated.
The formalisation initially follows the paper but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness leads to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulas in a branch of the tableau.
\end{itemize}

@@

All in all, the formalisation amounts to over 6000 lines of theory.
It is available online~\cite{formalisation}.

@@

Likewise, we fix the constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set!.\todo{Explain b'}
We overload this constant to return the set terms that are subterms of a set term, set atom, or formula, respectively. Lastly, we compute the subformulas of a formula with the function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set!.
The functions \lstinline!subterms! and \lstinline!subfms! are defined in the expected way (cf.\ \appendixname~\ref{appendix:subterms}).

\subsection{Semantics}
The original paper~\cite{new_fast_tableau} bases the semantics of \MLSS{} on the von Neumann hierarchy of sets $\mathcal{V}$.
While there is an entry~\cite{zfc_in_hol_afp} in the \emph{Archive of Formal Proofs} (AFP) that axiomatises this hierarchy,
we instead use the hierarchy of \emph{hereditarily finite sets} (HF sets) which fulfil all the same axioms as $\mathcal{V}$ -- that is, the axioms of ZF -- except for the axiom of infinity.
In particular, the membership relation is well-founded.
The HF sets are sufficient for our purposes as \MLSS{} is not powerful enough to force infinite models~\cite[Chapter~14.2]{set_theory}.\cite{Argue power: self-evident}
In contrast to $\mathcal{V}$, the HF sets are directly representable in Isabelle/HOL, and indeed, there is an AFP entry~\cite{hf_AFP} that formalises them.
The entry defines a type \lstinline!hf! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!HF :: hf set $\Rightarrow$ set! that converts a finite set of HF sets into an HF set.
\item The usual set operations such as equality ($=$), membership ($\hfmem$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: hf!.
\end{itemize}

Equipped with the above, we define the interpretation functions 
\begin{itemize}
  \item \lstinline!&\Ist& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_term $\Rightarrow$ hf! and 
  \item \lstinline!&\Isa& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_atom $\Rightarrow$ hf!
\end{itemize}
in the standard way, i.e.\ by mapping each syntactic construct to the corresponding operation on HF sets and interpreting variables with respect to a given valuation function \lstinline!M :: 'a $\Rightarrow$ hf!.
For the concrete definition we refer to formalisation (cf.\ \appendixname~\ref{appendix:semantics}).

@@

\section{A Tableau Calculus for MLSS}
We formalise the tableau calculus for \MLSS{} as described by \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}.
Inspired by the formalisation of a tableau calculus for hybrid logic by \citeauthor{hybrid_logic_afp}~\cite{hybrid_logic_afp}, we simply use a list to represent a branch of the tableau tree.
Note that formulas are added to the front of the list during branch expansion, so \lstinline!last b! for a branch \lstinline!b! is always the formula that we are trying to disprove with the tableau.
We sometimes call this formula the \emph{initial formula}.
\begin{lstlisting}
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}

@@

abbreviation bopen b $\equiv$ $\neg$ bclosed b
\end{lstlisting}
A tableau is called \emph{closed} if all of its branches are closed.\todo{What is a tableau?}

\subsection{Linear Expansion Rules}
\begin{table*}[t]
  \centering
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \caption{Linear expansion rules.\label[table]{tab:lexpands} We omit the rules for $\interS$ and $\diffS$ as they coincide with the original paper~\cite{new_fast_tableau} (cf.\ \appendixname~\ref{appendix:linear_rules}).}
  \begin{tabular}{ccc}
    \toprule
    Propositional Rules & {\hskip 6pt} & Rules for \lstinline!$\unionS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! \\
      \lstinline!@\fmNeg{(\fmOr{p}{q})}@! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@, @\fmNeg{q}@! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{p}@! & $\Longrightarrow$ & \lstinline!q! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{q}@! & $\Longrightarrow$ & \lstinline!p! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! \\
      \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p!\\[6ex]
    \end{tabular}
                        & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\in$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\[0.9em]
    \end{tabular}
    \\

@@

The calculus considers two kinds of branch expansion rules: \emph{linear} and \emph{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulas \lstinline!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\cref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to replace \lstinline!t! only if it occurs as a top-level set term of \lstinline!l!.
As an example, \lstinline!u! and \lstinline!s $\interS$ t! are top-level terms of \lstinline!u $\inS$ s $\interS$ t! but \lstinline!s! and \lstinline!t! are not.

A more crucial restriction of the linear rules is that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$! is
\lstinline!s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$!,
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
may only be used under the condition
\lstinline[breaklines]!t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b)!.
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulas in a branch in \cref{sec:correct}.

@@

Finally, we remark that the original paper~\cite{new_fast_tableau} is missing the last propositional rule dealing with double negation.
This rule is required for completeness, though, considering that the branch
\lstinline![&\fmNeg{\fmNeg{\fmNeg{p}}}&, p, &\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&]!
is saturated --- neither linear nor branching rules apply --- and open, but there clearly is no model for the initial formula \lstinline!&\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&!.\todo{Split into multiple sentences}

\subsection{Branching Rules\label[section]{sec:branching}}
\begin{table*}[t]
  \lstset{
    escapeinside={@}{@}
  }
  \centering
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \newcommand{\lpipe}{\rule[-0.4ex]{0.6pt}{2.3ex}}
  \caption{Branching expansion rules.\label[table]{tab:bexpands} We write \lstinline!$\phi$! for \lstinline!last b! here. The elided rules coincide with the original paper~\cite{new_fast_tableau} (cf.\ \appendixname~\ref{appendix:branching_rules}).}
  \begin{tabular}{>{\centering\arraybackslash}p{0.34\textwidth} >{\centering\arraybackslash}p{0.32\textwidth} >{\centering\arraybackslash}p{0.3\textwidth}}
    \toprule
    Rule & Precondition & Subsumption condition \\
    \midrule
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!p! & \lpipe & \lstinline!@\fmNeg{p}@! \\
    \end{tabular}
         & \lstinline!@\fmOr{p}{q}@ $\in$ set b! &
         \begin{tabular}{rl}
         & \lstinline!p $\in$ set b $\lor$! \\
         $\fmNegSymbol$ & \lstinline!p $\in$ set b! \\
         \end{tabular} \\

@@

The purpose of the function is to determine whether a given branch can be expanded to a closed tableau.
As stated before, we first use linear expansion rules in order to prevent premature branching;
to this end, we recursively expand the branch with linear rules until the branch is linearly saturated.
Note that we use Hilbert's $\varepsilon$-operator in the form of \lstinline!SOME! to choose some rule that actually adds new formulas to the branch. 
As soon as the branch is linearly saturated, we terminate if the branch is closed as shown in the second equation.
Otherwise, we choose an applicable branching rule and recursively check whether all newly created branches can be closed.
The final equation applies once no further branch expansion is possible, in which case we just test for closedness of the branch.

@@

\section{Soundness of the Calculus\label[section]{sec:sound}}
A tableau calculus is sound if for any closed tableau, the corresponding formula is unsatisfiable. 
We prove the following two properties to establish soundness:
\begin{enumerate*}[label={(\arabic*)}]
  \item A closed branch contains an unsatisfiable formula.
  \item The expansion rules maintain satisfiability.
\end{enumerate*}

@@

\section{Total Correctness of the Decision Procedure\label[section]{sec:correct}}
We first demonstrate the termination of the procedure for well-formed branches, i.e.\ every well-formed branch is in the domain of \lstinline!mlss_proc_branch!.
To this end, we derive an upper bound for the number of distinct formulas in a branch whose proof we omit here for brevity (cf.\ \appendixname~\ref{appendix:bound}).
We should point out that this bound is not to be construed as the complexity of the procedure as it may create exponentially many branches in general.
\begin{lstlisting}[label={lst:card_branch}]
lemma card_wf_branch_ub:
  assumes wf_branch b
  shows |set b| $\leq$ 2$\:$*$\,$|subfms (last b)| + 16$\:$*$\,$|subterms (last b)|$^\text{4}$
\end{lstlisting}
Remember that \lstinline!mlss_proc_branch! only applies a linear expansion rule to a branch if the application results in new formulas.   
Moreover, the subsumption conditions of the branching expansion rules ensure that each of the newly created branches contain new formulas.
Ultimately, we conclude that the procedure must terminate for well-formed branches because the number of formulas increases in each step but is also bounded. 
\begin{lstlisting}
lemma assumes wf_branch b shows mlss_proc_branch_dom b
\end{lstlisting}

@@

%\subsubsection{Acknowledgements}
%The author thanks Kevin Kappelmann for his comments on a draft version of the paper.

@@

%\clearpage\listoftodos
\appendix

\clearpage\chapter*{Appendix}
\todo[inline]{Remove appendix}
\section{Subterms and Subformulas\label{appendix:subterms}}

We introduce the function \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term! that is overloaded in its argument.
We first overload the function for set terms.
\begin{lstlisting}
fun subterms :: 'a pset_term $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\unionS$ t) = {s $\unionS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\interS$ t) = {s $\interS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\diffS$ t) = {s $\diffS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) = {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of its arguments.
\begin{lstlisting}
fun subterms :: 'a pset_atom $\Rightarrow$ 'a pset_term set where
  subterms (s $\inS$ t) = subterms s $\cup$ subterms t
| subterms (s $\eqS$ t) = subterms s $\cup$ subterms t
\end{lstlisting}
Finally, we lift this to formulas by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms :: 'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms $\phi$ $\equiv$ $\bigcup$(subterms ` atoms $\phi$)
\end{lstlisting}

We also consider the subformulas of a formula and define a function that computes them.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (&\fmAtom& a) = {&\fmAtom& a}
| subfms (&\fmAnd{p}{q}&) = {&\fmAnd{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmOr{p}{q}&) = {&\fmOr{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmNeg{q}&) = {&\fmNeg{q}&} $\cup$ subfms q
\end{lstlisting}

Finally, we lift the functions \lstinline!vars! and \lstinline!subterms! to branches.
\begin{lstlisting}
definition vars :: 'a branch $\Rightarrow$ 'a set 
  vars b $\equiv$ $\bigcup$(vars ` set b)

definition subterms :: 'a branch $\Rightarrow$ 'a pset_term set where
  subterms b $\equiv$ $\bigcup$(subterms ` set b)
\end{lstlisting}

\clearpage

\section{Definition of the Interpretation Functions\label{appendix:semantics}}

We define the interpretation function \lstinline!&\Ist&! for set atoms.
\begin{lstlisting}
fun &\Ist& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_term $\Rightarrow$ hf where
  &\Ist& M $\emptyset$ = 0
| &\Ist& M (Var x) = M x
| &\Ist& M (Single s) = HF {&\Ist& M s}
| &\Ist& M (s $\unionS$ t) = &\Ist& M s $\sqcup$ &\Ist& M t 
| &\Ist& M (s $\interS$ t) = &\Ist& M s $\sqcap$ &\Ist& M t 
| &\Ist& M (s $\diffS$ t) = &\Ist& M s $-$ &\Ist& M t 
\end{lstlisting}
The interpretation function \lstinline!I$_\text{sa}$! for set atoms is straightforward as well.
\begin{lstlisting}
fun &\Isa& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_atom $\Rightarrow$ bool where
  &\Isa& M (s $\inS$ t) $\longleftrightarrow$ &\Ist& M s $\hfmem$ &\Ist& M t
| &\Isa& M (s $\eqS$ t) $\longleftrightarrow$ &\Ist& M s = &\Ist& M t
\end{lstlisting}

\section{Upper-Bounding the Cardinality of a Branch\label{appendix:bound}}
The argument below argues about the \emph{literals} of a branch, which are defined as follows.
\begin{lstlisting}
fun is_literal :: 'a fm $\Rightarrow$ bool where
  is_literal (&\fmAtom& _) = True | is_literal (&\fmNeg{(\fmAtom{} \_)}&) = True
| is_literal _ = False
\end{lstlisting}

We still need to give a termination proof to ascertain total correctness.
The core of our argument is that the number of distinct formulas that can be derived with the expansion rules is finite.
In particular, we claim that the following upper bound holds.
\begin{lstlisting}
  &\phantom{$\leq$}& |set b|
  &\textover{$=$}{$\leq$}& |{$\psi$ $\in$ set b. $\neg$ is_literal $\psi$}| +
  &\phantom{$\leq$}& |{$\psi$ $\in$ set b. is_literal $\psi$}|
  $\leq$ 2 * |subfms (last b)| + |{$\psi$ $\in$ set b. is_literal $\psi$}|
  $\leq$ 2 * |subfms (last b)| + 16 * |subterms (last b)|$^4$
\end{lstlisting}
We treat both inequalities separately, starting with the first one whose justification hinges on an invariant: 
every new composite subformula that is introduced by an expansion rule is the negation of a subformula that already occurs in \lstinline!last b!.
Written as a formula, we have
\begin{lstlisting}
  &\phantom{$\subseteq$}& {$\psi$ $\in$ set b. $\neg$ is_literal $\psi$}
  $\subseteq$ subfms (last b) $\cup$ $\fmNegSymbol$ ` subfms (last b).
\end{lstlisting}
Confirming that this is an invariant is an easy exercise of going through each of the propositional rules in \cref{tab:lexpands} and the first two branching rules in \cref{tab:bexpands}.
Thus, we have
\begin{lstlisting}
  &\phantom{$\leq$}& |{$\psi$ $\in$ set b. $\neg$ is_literal $\psi$}|
  $\leq$ |subfms (last b)| + |$\fmNegSymbol$ ` subfms (last b)|
  $\leq$ 2 * |subfms (last b)|
\end{lstlisting}
which is the first inequality we asserted.

>>>>

@@

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is a quantifier-free fragment of set theory.
  We formalise its syntax and semantics as well as a sound and complete tableau calculus for it.
  We also provide an executable specification of a decision procedure that exhaustively applies the rules of the calculus and prove its termination.
  Furthermore, we extend the calculus with a lightweight type system that paves the way for an integration of the procedure into Isabelle/HOL.
  \keywords{Decision procedures \and Semantic tableaux \and Interactive theorem proving \and Set theory}
\end{abstract}

@@

\subsection{Contributions}
We present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that exhaustively applies the rules of the calculus.
To obtain total correctness of the procedure, we prove its termination.
Additionally, we naively refine the abstract to an executable specification from which we can generate code.
The formalisation initially follows the paper but offers a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness leads to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulas in a branch of the tableau.
\end{itemize}

@@

All in all, the formalisation amounts to over 6000 lines of theory.
It is part of the \emph{Archive of Formal Proofs} (AFP)~\cite{formalisation}.
The entry provides an overview theory \href{https://www.isa-afp.org/theories/mlss_decision_proc/#MLSS_Proc_All}{\texttt{MLSS\_Proc\_All.thy}} that highlights the (mostly syntactic) differences between paper and formalisation and references the constants and theorems that are introduced in this paper.

@@

Likewise, we fix the constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set! that is polymorphic in its argument type \lstinline!'b!.
We overload this constant to return the set terms that are subterms of a set term, set atom, or formula, respectively.
Lastly, we introduce the function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set! that computes the subformulas of a formula.
The functions \lstinline!subterms! and \lstinline!subfms! are implemented in the expected way.

\subsection{Semantics}
The original paper~\cite{new_fast_tableau} bases the semantics of \MLSS{} on the von Neumann hierarchy of sets $\mathcal{V}$.
We instead use the hierarchy of \emph{hereditarily finite sets} (HF sets) which fulfil all the same axioms as $\mathcal{V}$ -- that is, the axioms of ZF -- except for the axiom of infinity.
In particular, the membership relation is well-founded.
The HF sets, as we will see, are sufficient to construct a model for any satisfiable \MLSS{} formula.
In contrast to $\mathcal{V}$, the HF sets are directly representable in Isabelle/HOL, and indeed, there is an AFP entry~\cite{hf_AFP} that formalises them.
The entry defines a type \lstinline!hf! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!HF :: hf set $\Rightarrow$ set! that converts a finite set of HF sets into an HF set.
\item The usual set operations such as equality ($=$), membership ($\hfmem$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: hf!.
\end{itemize}

Equipped with the above, we define the interpretation functions 
\begin{itemize}
  \item \lstinline!&\Ist& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_term $\Rightarrow$ hf! and 
  \item \lstinline!&\Isa& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_atom $\Rightarrow$ hf!
\end{itemize}
in the standard way, i.e.\ by mapping each syntactic construct to the corresponding operation on HF sets and interpreting variables with respect to a given valuation function \lstinline!M :: 'a $\Rightarrow$ hf!.
For the concrete definition we refer to formalisation.

@@

\section{A Tableau Calculus for MLSS}
We formalise the tableau calculus for \MLSS{} as described by \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}.
Inspired by the formalisation of a tableau calculus for hybrid logic by \citeauthor{hybrid_logic_afp}~\cite{hybrid_logic_afp}, we simply use lists to represent the branches of the tableau tree.
Note that formulas are added to the front of the list during branch expansion, so \lstinline!last b! for a branch \lstinline!b! is always the formula that we are trying to disprove with the tableau.
We sometimes call this formula the \emph{initial formula}.
\begin{lstlisting}
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}

@@

abbreviation bopen b $\equiv$ $\neg$ bclosed b
\end{lstlisting}
A tableau is called \emph{closed} if all of its branches are closed.

\subsection{Linear Expansion Rules}
\begin{table*}[t]
  \centering
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \caption{Linear expansion rules.\label[table]{tab:lexpands} All rules except the double negation rule coincide with the original paper~\cite{new_fast_tableau}. For brevity, we omit the rules for $\interS$ and $\diffS$.}
  \begin{tabular}{ccc}
    \toprule
    Propositional Rules & {\hskip 6pt} & Rules for \lstinline!$\unionS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! \\
      \lstinline!@\fmNeg{(\fmOr{p}{q})}@! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@, @\fmNeg{q}@! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{p}@! & $\Longrightarrow$ & \lstinline!q! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{q}@! & $\Longrightarrow$ & \lstinline!p! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! \\
      \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p!\\[6ex]
    \end{tabular}
                        & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\in$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\[0.9em]
    \end{tabular}
    \\

@@

The calculus considers two kinds of branch expansion rules: \emph{linear} and \emph{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulas \lstinline!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\cref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to the top-level set terms of \lstinline!l!, i.e.\ the set terms that occur directly under one of the atom constructors \lstinline!$\inS$! or \lstinline!$\eqS$!;
for example, given the literal
\begin{lstlisting}
  l = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&
\end{lstlisting}
we have
\begin{lstlisting}
  &\phantom{=}& (&\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&){t/s $\unionS$ u}
  = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ t)}&.
\end{lstlisting}
A more crucial restriction of the linear rules is that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$! is
\begin{lstlisting}
  s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$,
\end{lstlisting}
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
and may only be used under the condition
\lstinline[breaklines]!t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b)!.
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulas in a branch in \cref{sec:correct}.

@@

Finally, we remark that the original paper~\cite{new_fast_tableau} is missing the last propositional rule dealing with double negation.
This rule is required for completeness, though, considering that the branch
\lstinline![&\fmNeg{\fmNeg{\fmNeg{p}}}&, p, &\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&]!
is saturated --- neither linear nor branching rules apply --- and open, but there clearly is no model for the initial formula \lstinline!&\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&!.

\subsection{Branching Rules\label[section]{sec:branching}}
\begin{table*}[t]
  \lstset{
    escapeinside={@}{@}
  }
  \centering
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \newcommand{\lpipe}{\rule[-0.4ex]{0.6pt}{2.3ex}}
  \caption{Branching expansion rules.\label[table]{tab:bexpands} We write \lstinline!$\phi$! for \lstinline!last b! here. All rules coincide with the original paper~\cite{new_fast_tableau} so we only show an illustrative subset.}
  \begin{tabular}{>{\centering\arraybackslash}p{0.34\textwidth} >{\centering\arraybackslash}p{0.32\textwidth} >{\centering\arraybackslash}p{0.3\textwidth}}
    \toprule
    Rule & Precondition & Subsumption condition \\
    \midrule
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!p! & \lpipe & \lstinline!@\fmNeg{p}@! \\
    \end{tabular}
         & \lstinline!@\fmOr{p}{q}@ $\in$ set b! &
         \begin{tabular}{rl}
         & \lstinline!p $\in$ set b $\lor$! \\
         $\fmNegSymbol$ & \lstinline!p $\in$ set b! \\
         \end{tabular} \\

@@

The purpose of the function is to determine whether a given branch can be expanded to a closed tableau.
As stated before, we first use linear expansion rules in order to prevent premature branching;
to this end, we recursively expand the branch with linear rules until the branch is linearly saturated.
Note that we use Hilbert's $\varepsilon$-operator in the form of \lstinline!SOME!\footnote{In the formalisation, the function \lstinline!mlss_proc_branch! is actually parametrised by choice functions to allow for refinement.} to choose some rule that actually adds new formulas to the branch. 
As soon as the branch is linearly saturated, we terminate if the branch is closed as shown in the second equation.
Otherwise, we choose an applicable branching rule and recursively check whether all newly created branches can be closed.
The final equation applies once no further branch expansion is possible, in which case we just test for closedness of the branch.

@@

\section{Soundness of the Calculus\label[section]{sec:sound}}
A tableau calculus is sound if for any closed tableau, the corresponding formula is unsatisfiable. 
We prove the following two properties to establish soundness:
\begin{enumerate*}[label={(\arabic*)}]
  \item It is not possible to satisfy all formulas in a closed branch at once. 
  \item The expansion rules maintain satisfiability.
\end{enumerate*}

@@

\section{Total Correctness of the Decision Procedure\label[section]{sec:correct}}
We first demonstrate the termination of the procedure for well-formed branches, i.e.\ every well-formed branch is in the domain of \lstinline!mlss_proc_branch!.
To this end, we derive an upper bound for the number of distinct formulas in a branch whose proof we omit here for brevity.
We should point out that this bound is not to be construed as the complexity of the procedure as it may create exponentially many branches in general.
\begin{lstlisting}[label={lst:card_branch}]
lemma card_wf_branch_ub:
  assumes wf_branch b
  shows |set b| $\leq$ 2$\:$*$\,$|subfms (last b)| + 16$\:$*$\,$|subterms (last b)|$^\text{4}$
\end{lstlisting}
Remember that \lstinline!mlss_proc_branch! only applies a linear expansion rule to a branch if the application results in new formulas.   
Moreover, the subsumption conditions of the branching expansion rules ensure that each of the newly created branches contain new formulas.
Ultimately, we conclude that the procedure must terminate for well-formed branches because the number of formulas increases in each step but is also bounded. 
\begin{lstlisting}
lemma assumes wf_branch b shows mlss_proc_branch_dom b
\end{lstlisting}

@@

\subsubsection{Acknowledgements}
The author thanks Kevin Kappelmann, Tobias Nipkow, and the anonymous referees for their comments on a draft version of this paper.

@@

\clearpage\listoftodos
